Enhancing Multiple Myeloma Cell Segmentation through Physical Image Transformations

### 2023-05-02 paper/paper.tex
Updated the paper

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{makecell}
\usepackage{algorithm}
\usepackage{algpseudocode}%
\usepackage{bm}
\usepackage{amsmath}
\usepackage{makecell}

@@

Advanced digital image processing techniques, such as object detection and segmentation, can help overcome these limitations by automating the abnormal plasma cell detection process. By combining these techniques with expert pathologist decision-making, it may be possible to reduce diagnosis time and alleviate the workload of pathologists. The primary objective of developing an automated diagnostic tool for Multiple Myeloma is to perform accurate segmentation of cells, which is a challenging task due to the variety of possible scenarios. The SegPC-2021 challenge aimed to address this issue by providing bone marrow aspirate slide images from Multiple Myeloma patients. Segmentation of plasma cells in these images is complicated due to several factors, such as the presence of cell clusters, cells with varying nucleus and cytoplasm sizes, and difficulty in distinguishing cells that are touching. Additionally, the staining process used in preparing the slides may not be perfect, resulting in cells that are not stained or cells with cytoplasm that blends with the background color. The task becomes even more challenging due to the use of computer-generated labels, which may contain inconsistencies and noise.

Research on automated cell segmentation in microscopic images has been ongoing for many years. However, recent advancements in deep learning have led to significant breakthroughs in this field. U-Net, a fully convolutional neural network architecture, has demonstrated remarkable results in cell segmentation and tracking challenges. Similarly, Mask R-CNN, a universal approach to instance segmentation, has been adapted to different domains, including cell nucleus segmentation. Instead of purely optimizing deep learning model architectures, it is hypothesized that optimizing imaging systems could improve the performance of instance segmentation. However, constructing and implementing different imaging systems can be both costly and time-consuming. Simulating various physical imaging systems and identifying the best performing one, on the other hand, is a more feasible approach. The purpose of this study is to devise an improved method for the detection of Multiple Myeloma cells in microscopic images by utilizing existing object detection and instance segmentation algorithms and augmenting them with various image transformation techniques, thereby simulating various physical imaging systems.

@@

In Gupta A et al [1], the paper discussed multiple approaches to tackle the challenge on segmentation of Multiple Myeloma cells on microscopic images. The top solution combined three instance segmentation models with three convolutional feature extractors, extensive image augmentation, and a custom model ensemble strategy. It utilized a post-processing step to pair corresponding instances of nuclei and cytoplasm together, using an auxiliary whole-cell class. The second-placed solution used a two-stage approach where the first stage consisted of extraction of possible nucleus instances using a U-Net and the second stage relied on dilated convolution and attention mechanism using Attention Deeplabv3+ for cytoplasm segmentation.

Paing MP et al [2] explored instance segmentation of Multiple Myeloma cells using Mask Region-based Convolutional Neural Network (R-CNN) models. They applied deep-wise augmentation, a deep learning-based data augmentation method to increase the performance of their models. However, none of these previous works incorporated physical imaging techniques to enhance the identification of Multiple Myeloma cells.

@@

To use the dataset in our experiments, we first resized the images to the dimensions of 1080x1440 pixels since the desired outputs were expected in this size and then converted the given ground truth image masks into Common Objects in Context (COCO) annotation format. The COCO annotation format is a JSON file that contains the image ID, image size, and the bounding box and segmentation mask coordinates of the nucleus and cytoplasm of the cell. The bounding box coordinates are provided in the form of the top-left and bottom-right coordinates of the bounding box. The segmentation mask coordinates represent the polygonized contour of the Multiple Myeloma cells. The COCO annotation format is used in the Detectron2 library for object detection and instance segmentation. The Detectron2 library was used for training and evaluating the Mask R-CNN model in our experiments.

@@

@@

@@

\subsubsection{Blur Filters}

@@

@@

A Mask R-CNN model was used to perform instance segmentation of Multiple Myeloma cell nucleus and cytoplasm. The model was implemented using the PyTorch deep learning framework and the Detectron2 library. It consisted of a backbone network, a Region Proposal Network (RPN) and two Mask Heads. The backbone network was used to extract features from the input image, which were then fed to the RPN to generate region proposals. The Mask Heads were used to predict the nucleus segmentation and cytoplasm segmentation masks for each region proposal. EfficientNet B5 was chosen as the backbone network. In the case of illumination simulation, an additional non-negative convolutional layer was added to convolve the illuminated images prior to being passed to EfficientNet. The selection of this non-negative attribute was a conscious choice as it is illogical for illumination weights to bear negative values in real life. The outputs of the model were the predicted probability scores of nucleus segmentation mask and cytoplasm segmentation mask. The model was trained for a total of 3725 steps with a learning rate of 0.0025 and a Warmup Cosine Learning Rate scheduler. Training was done on a single NVIDIA RTXA6000 GPU with 48GB of memory. The model was trained on the train set and evaluated on the validation set. The model with the best validation loss was selected for submission to the SEG-PC 2021 challenge website to obtain the Mean IOU score on the test set. A total of 11 models were trained, among which were 10 that incorporated one of the aforementioned physical image transformations and 1 that did not have any transformation.

@@

\bibitem{} Gupta A, Shiv Gehlot, Goswami S, et al. SegPC-2021: A challenge \& dataset on segmentation of Multiple Myeloma plasma cells from microscopic images. Medical Image Analysis. 2022;83:102677-102677. doi:https://doi.org/10.1016/j.media.2022.102677

\bibitem{} Paing MP, Sento A, Bui TH, Pintavirooj C. Instance Segmentation of Multiple Myeloma Cells Using Deep-Wise Data Augmentation and Mask R-CNN. Entropy. 2022;24(1):134. doi:https://doi.org/10.3390/e24010134

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{makecell}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{makecell}
\usepackage{subcaption}

@@

Advanced digital image processing techniques, such as object detection and segmentation, can help overcome these limitations by automating the abnormal plasma cell detection process. By combining these techniques with expert pathologist decision-making, it may be possible to reduce diagnosis time and alleviate the workload of pathologists. The primary objective of developing an automated diagnostic tool for Multiple Myeloma is to perform accurate segmentation of cells, which is a challenging task due to the variety of possible scenarios. The SegPC-2021 Challenge [1] aimed to address this issue by providing bone marrow aspirate slide images from Multiple Myeloma patients. Segmentation of plasma cells in these images is complicated due to several factors, such as the presence of cell clusters, cells with varying nucleus and cytoplasm sizes, and difficulty in distinguishing cells that are touching. Additionally, the staining process used in preparing the slides may not be perfect, resulting in cells that are not stained or cells with cytoplasm that blends with the background color. The task becomes even more challenging due to the use of computer-generated labels, which may contain inconsistencies and noise.

Research on automated cell segmentation in microscopic images has been ongoing for many years. However, recent advancements in deep learning have led to significant breakthroughs in this field. U-Net, a fully convolutional neural network architecture, has demonstrated remarkable results in cell segmentation and tracking challenges. Similarly, Mask R-CNN [2], a universal approach to instance segmentation, has been adapted to different domains, including cell nucleus segmentation. Instead of purely optimizing deep learning model architectures, it is hypothesized that optimizing imaging systems could improve the performance of instance segmentation. However, constructing and implementing different imaging systems can be both costly and time-consuming. Simulating various physical imaging systems and identifying the best performing one, on the other hand, is a more feasible approach. The purpose of this study is to devise an improved method for the detection of Multiple Myeloma cells in microscopic images by utilizing existing object detection and instance segmentation algorithms and augmenting them with various image transformation techniques, thereby simulating various physical imaging systems.

@@

In Gupta A et al [3], the paper discussed multiple approaches to tackle the challenge on segmentation of Multiple Myeloma cells on microscopic images. The top solution combined three instance segmentation models with three convolutional feature extractors, extensive image augmentation, and a custom model ensemble strategy. It utilized a post-processing step to pair corresponding instances of nuclei and cytoplasm together, using an auxiliary whole-cell class. The second-placed solution used a two-stage approach where the first stage consisted of extraction of possible nucleus instances using a U-Net and the second stage relied on dilated convolution and attention mechanism using Attention Deeplabv3+ for cytoplasm segmentation.

Paing MP et al [4] explored instance segmentation of Multiple Myeloma cells using Mask Region-based Convolutional Neural Network (R-CNN) models. They applied deep-wise augmentation, a deep learning-based data augmentation method to increase the performance of their models. However, none of these previous works incorporated physical imaging techniques to enhance the identification of Multiple Myeloma cells.

@@

To use the dataset in our experiments, we first resized the images to the dimensions of 1080x1440 pixels since the desired outputs were expected in this size and then converted the given ground truth image masks into Common Objects in Context (COCO) annotation format. The COCO annotation format is a JSON file that contains the image ID, image size, and the bounding box and segmentation mask coordinates of the nucleus and cytoplasm of the cell. The bounding box coordinates are provided in the form of the top-left and bottom-right coordinates of the bounding box. The segmentation mask coordinates represent the polygonized contour of the Multiple Myeloma cells. The COCO annotation format is used in the Detectron2 [5] library for object detection and instance segmentation. The Detectron2 library was used for training and evaluating the Mask R-CNN model in our experiments.

@@

\begin{figure}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=.4\linewidth]{1.png}
    \captionof{figure}{Original Image}
  \end{minipage}%
  \begin{minipage}{.4\textwidth}
    \centering
    \includegraphics[width=.6\linewidth]{2.jpg}
    \captionof{figure}{Simulation of 12 light waves hitting a sample from different angles
    }
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=.4\linewidth]{3.png}
    \captionof{figure}{Image with contrast normalization}
  \end{minipage}%
  \begin{minipage}{.4\textwidth}
    \centering
    \includegraphics[width=.6\linewidth]{4.png}
    \captionof{figure}{Image after Dilation operation}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=.4\linewidth]{5.png}
    \captionof{figure}{Image after Erosion operation}
  \end{minipage}%
  \begin{minipage}{.4\textwidth}
    \centering
    \includegraphics[width=.6\linewidth]{6.png}
    \captionof{figure}{Image with Sobel Gradient filter}
  \end{minipage}
\end{figure}
\subsubsection{Blur Filters}

@@

\begin{figure}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=1\linewidth]{7.png}
    \captionof{figure}{Image after Erosion operation}
  \end{minipage}%
\end{figure}

\begin{table}[b]
  \caption{Performance comparison of the 11 models on the validation and test sets}
  \label{sample-table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    \multicolumn{2}{c}{ }\\
    {\bf Image} & {\bf Nucleus Segmentation} & {\bf Cytoplasm Segmentation} & {\bf Test} \\
    {\bf Transformation} & {\bf Validation AP} & {\bf Validation AP} & {\bf Mean IOU}   \\
    \midrule
    None & 80.39 & 44.34 & 0.85 \\
    \midrule
    Illumination Simulation & 77.64 & 41.46 & 0.84 \\
    \midrule
    Contrast Normalization & 78.21 & 44.05 & 0.84 \\
    \midrule
    Erosion & 78.76 & 43.83 & 0.84 \\
    \midrule
    Dilation & {\bf 82.00} & 43.79 & 0.85 \\
    \midrule
    Sobel Gradient Filter & 74.13 & 38.53 & 0.83 \\
    \midrule
    Red Channel Enhancement & 79.87 & 42.97 & 0.85 \\
    \midrule
    Green Channel Enhancement & 79.91 & 42.32 & 0.86 \\
    \midrule
    Blue Channel Enhancement & 79.04 & 43.51 & 0.85 \\
    \midrule
    Gaussian Blur Filter & 81.74 & 43.95 & 0.84 \\
    \midrule
    Median Blur Filter & 81.52 & {\bf 46.03} & {\bf 0.87} \\
   \bottomrule
 \end{tabular}
\end{table}

@@

A Mask R-CNN model was used to perform instance segmentation of Multiple Myeloma cell nucleus and cytoplasm. The model was implemented using the PyTorch deep learning framework and the Detectron2 library. It consisted of a backbone network, a Region Proposal Network (RPN) and two Mask Heads. The backbone network was used to extract features from the input image, which were then fed to the RPN to generate region proposals. The Mask Heads were used to predict the nucleus segmentation and cytoplasm segmentation masks for each region proposal. EfficientNet B5 [6] was chosen as the backbone network. In the case of illumination simulation, an additional non-negative convolutional layer was added to convolve the illuminated images prior to being passed to EfficientNet. The selection of this non-negative attribute was a conscious choice as it is illogical for illumination weights to bear negative values in real life. The outputs of the model were the predicted probability scores of nucleus segmentation mask and cytoplasm segmentation mask. The model was trained for a total of 3725 steps with a learning rate of 0.0025 and a Warmup Cosine Learning Rate scheduler. Training was done on a single NVIDIA RTXA6000 GPU with 48GB of memory. The model was trained on the train set and evaluated on the validation set. The model with the best validation loss was selected for submission to the SEG-PC 2021 challenge website to obtain the Mean IOU score on the test set. A total of 11 models were trained, among which were 10 that incorporated one of the aforementioned physical image transformations and 1 that did not have any transformation.

@@

  \bibitem{} SegPC-2021 - Grand Challenge. grand-challenge.org. Accessed April 19, 2023. https://segpc-2021.grand-challenge.org
  \bibitem{} He K, Gkioxari G, Dollár P, Girshick R. Mask R-CNN. arXiv.org. Published 2017. doi: https://doi.org/10.48550/arXiv.1703.06870
  \bibitem{} Gupta A, Shiv Gehlot, Goswami S, et al. SegPC-2021: A challenge \& dataset on segmentation of Multiple Myeloma plasma cells from microscopic images. Medical Image Analysis. 2022;83:102677-102677. doi:https://doi.org/10.1016/j.media.2022.102677
  \bibitem{} Paing MP, Sento A, Bui TH, Pintavirooj C. Instance Segmentation of Multiple Myeloma Cells Using Deep-Wise Data Augmentation and Mask R-CNN. Entropy. 2022;24(1):134. doi:https://doi.org/10.3390/e24010134
  \bibitem{} Wu Y, Kirillov A, Massa F, Lo W, Girshick R. Detectron2. Published 2019. https://github.com/facebookresearch/detectron2
  \bibitem{} Tan M, Le QV. EfficientNet: Rethinking Model Scaling for Convolutional Neural Networks. arXiv.org. Published 2019. doi: https://doi.org/10.48550/arXiv.1905.11946
